High-dimensional Asymptotics of Transfer Learning for Linear Regression

### 2021-06-12 revision_journal/backup/Jan 31 2021/manuscript.tex
revision wip

\begin{abstract}
	{\cor rewrite} Hard parameter sharing for multi-task learning is widely used in empirical research despite the fact that its generalization properties have not been well established in many cases. This paper studies its generalization properties in a fundamental setting: How does hard parameter sharing work given multiple linear regression tasks? We develop new techniques and establish a number of new results in the high-dimensional setting, where the sample size and feature dimension increase at a fixed ratio. First, we show a sharp bias-variance decomposition of hard parameter sharing, given multiple tasks with the same features. Second, we characterize the asymptotic bias-variance limit for two tasks, even when they have arbitrarily different sample size ratios and covariate shifts. We also demonstrate that these limiting estimates for the empirical loss are incredibly accurate in moderate dimensions. Finally, we explain an intriguing phenomenon where increasing one task's sample size helps another task initially by reducing variance but hurts eventually due to increasing bias. This suggests progressively adding data for optimizing hard parameter sharing, and we validate its efficiency in text classification tasks.
\end{abstract}

@@

{\cor rewrite} Hard parameter sharing (HPS) for multi-task learning is widely used in empirical research and goes back to the seminal work of \cite{C97}.
Recent work has revived interests in this approach because it improves performance and reduces the cost of collecting labeled data \cite{R17}.
It is generally applied by sharing the feature layers between all tasks while keeping an output layer for every task.
Often, hard parameter sharing offers two critical advantages if successfully applied.
First, it reduces model parameters since all tasks use the same feature space.
Second, it reduces the amount of labeled data needed from each task by augmenting the entire training dataset.

@@

\FY{Introduce some popular transfer learning estimators: total loss estimator, and model averaging estimator.}

We study the standard hard parameter sharing architecture: a shared feature representation layer $B\in\real^{p}$ for all datasets and a separate output layer $A_i \in \real$ for every dataset $i$. Then we study the following minimization problem:
\begin{align}\label{eq_tsl}
			f(A, B) = \norm{X^{(1)} B A_1 - Y^{(1)}}^2 + \norm{X^{(2)} B A_2 - Y^{(2)}}^2,
\end{align}
where we abbreviate $A = [A_1, A_2]$. Let $(\hat{A}, \hat{B})$ be the minimizer of $f(A, B)$. We define the hard parameter sharing (HPS) estimator for task $i$ as 
\be\label{def_HPS}\hat{\beta}_i^{\MTL} = \hat{B} \hat{A}_i,\quad i=1,2.\ee

@@

In this paper, we study the out-of-sample predication loss (test error) of $\hat{\beta}_2^{\MTL} (\hat a)$. Consider a test data point $(x,y)$ generated from the same model as task 2: $y= x^\top \beta^{(2)} + \ve$, where $x \in \R^p$ and $\ve\in \R$ are independent of $X^{(1)}$, $X^{(2)}$, $\varepsilon^{(1)}$ and $\varepsilon^{(2)}$, and only $x$ is observable. We want to use $x^\top \hat{\beta}_2^{\MTL} (\hat a) $ to predict $y$, and we measure the predication loss using the mean squared error
$$\exarg{x}{\left\|y-x^\top \hat{\beta}_2^{\MTL} (\hat a) \right\|^2}= \left\|(\Sigma^{(2)})^{1/2} \left(\hat{\beta}_2^{\MTL} (\hat a)  - \beta^{(2)}\right)\right\|^2 + \sigma^2.  $$ 
Since $\sigma^2$ is a constant that does not depend on the model, we ignore it and define the predication loss as
\be\label{HPS_loss}
L(\hat{\beta}_2^{\MTL}(\hat a)):= \left\|(\Sigma^{(2)})^{1/2} \left(\hat{\beta}_2^{\MTL} (\hat a)  - \beta^{(2)}\right)\right\|^2 .
 \ee
We will compare it to the out-of-sample predication loss $L(\hat{\beta}_2^{\STL})$ of the single-task OLS estimator 
$$\hat \beta^{\STL}= [(X^{(2)})^\top X^{(2)}]^{-1}(X^{(2)})^\top Y^{(2)}$$ as a baseline.

@@

 \begin{lemma}\label{lem_HPS_loss}
 Under Assumption \ref{assm_big1}, for any small constant $\e>0$ and large constant $C>0$, there exists a high probability event $\Xi$, on which the following estimates hold uniformly in $a\in \R$: 
\begin{align}
L(\hat{\beta}_2^{\MTL}(a)) &=\left[ 1+ \OO(p^{-1/2+c})\right]\cdot \left[L_{\bias} (a) + L_{\var}(a)\right] \nonumber\\
&+ \OO \left[ p^{-C}\left( \|\beta^{(1)} \|^2  +  \|\beta^{(2)} \|^2\right)   \right], \label{L_HPS_simple}
\end{align}
and 
\begin{align}
L(\hat{\beta}_2^{\STL} ) &= \left[ 1+ \OO(p^{-1/2+c})\right] \cdot \sigma^2  \bigtr{\Sigma^{(2)} [(X^{(2)})^\top X^{(2)}]^{-1}  }  . \label{L_STL_simple}
\end{align}
%Moreover, there exists a constant $C_0>0$ such that with high probability,
%\be\label{loss_large}
%g(a) > g(0) \quad \text{for all $a$ such that }|a|\ge C_0.
%\ee
%with high probability for any small constant $\e>0$. 
\end{lemma}

@@

\begin{lemma}[Theorem 2.4 of \cite{isotropic} and Theorem 3.14 of \cite{DY}]\label{fact_tr}
	%Let $X  \in \real^{n\times p}$ be a random matrix that satisfies Assumption \ref{assume_rm}.
	%Let $\Sigma\in\real^{p\times p}$ denote the population covariance matrix of $X$.
	Under Assumption \ref{assm_big1}, we have that
		\[ \bigtr{\Sigma^{(2)} \frac1{(X^{(2)})^\top X^{(2)} }  } = \bigtr{ \frac1{(Z^{(2)})^\top Z^{(2)} }  } = \frac{p}{n_2 - p} \cdot \left[1+ \OO\left( \frac{p^{\e}}{\sqrt{np}} \right)\right] \]
		with high probability for any small constant $\e>0$.
\end{lemma}
 %The above result has a long history in random matrix theory.
If the entries of $Z^{(2)}$ are i.i.d. Gaussian, then this result follows from the classical result for the mean of inverse Wishart distribution \cite{anderson1958introduction}. For a general non-Gaussian $Z^{(2)}$, this result can be obtained using the well-known Stieltjes transform method (cf. Lemma 3.11 of \cite{bai2009spectral}). Here we have presented the results from \cite{isotropic,DY}, which give an almost sharp convergence rate. With Lemma \ref{fact_tr}, we obtain that w.h.p.,
\begin{align}
L(\hat{\beta}_2^{\STL} ) &= \left[ 1+ \OO(p^{-1/2+c})\right] \cdot \frac{p \sigma^2 }{n_2-p}   . \label{L_STL_simple01}
\end{align}

On the other hand, much less is known about the predication loss of the HPS estimator. In particular, its exact dependence on the model parameters, including the model bias, noise variance, sample sizes, and population covariance matrices, is not well-understood so far. \FY{What is known so far about this topic?} In the rest of this paper, we give a rigorous analysis of the bias term \eqref{Lbias} and variance term \eqref{Lvar} in three different settings of increasing complexity: (i) same covariates $X^{(1)}=X^{(2)}$; (ii) independent $X^{(1)}$ and $X^{(2)}$ with different sample sizes and same population covariance matrices $\Sigma^{(1)}=\Sigma^{(2)}$; (iii) independent $X^{(1)}$ and $X^{(2)}$ with different sample sizes and different population covariance matrices. In each case, we provide the exact asymptotic bias and variance limits, together with almost sharp convergence rates.  
In particular, we will use our results to analyze the effect of the following factors: \emph{bias-variance tradeoff}, \emph{sample sizes}, and \emph{covariate shift}.

{\cor discuss about the bias-variance trade-off}
%However, to the best of our knowledge, the asymptotic limit of $L(\hat{\beta}_2^{\MTL}(a))$ has never been calculated exactly in any high-dimensional setting.

\section{Predication loss for HPS estimator}

@@

 In this section, we consider the setting where $X^{(1)}$ and $X^{(2)}$ are independent and have the same population covariance matrices $\Sigma^{(1)}=\Sigma^{(2)}$. However, the two tasks can have different sample sizes $n_1\ne n_2$. In this case, we can obtain the exact asymptotic limits of the bias term \eqref{Lbias} and the variance term \eqref{Lvar}. We will use these results to illustrate the effect of the sample sizes on the predication loss of the $\MTL$ estimator.

@@

Under Assumption \ref{assm_big1} holds, suppose that $\Sigma^{(1)}=\Sigma^{(2)}$ and the entries of $Z^{(1)}$ and $Z^{(2)}$ are i.i.d. Gaussian random variables. Then for any small constant $\e>0$ and large constant $C>0$, there exists a high probability event $\Xi$, on which the following estimates hold uniformly in all $a\in \R$: 

@@

L_{\var}(a)= \left[ \cal L_1(a)+ \OO\left(\frac{p^{c}}{n_1}\right)\right] \cdot   \sigma^2 ,

@@

L_{\bias}(a)&= \left[ \cal L_2(a)+  \OO\left(p^{-1/2+c}\right)\right]   \|\beta^{(1)}-a\beta^{(2)}\|^2 \nonumber\\

@@

\iffalse
 We need to state a result for Gaussian matrix ...... Consider 
$$f (\al,n_1,n_2)= \frac1p\tr\left[\frac{1}{ (X_1^\top X_1 + \al \cdot X_2^\top X_2)^2} (X_1^\top X_1)^2\right].$$
In our case, we have $\al=1$, but we can handle more general $\al$. We introduce two parameters:
$$a= \al \frac{n_2}{n_1} \left( \frac{p}{n_1} + \frac{p}{n_2}- \frac{p}{n_1 }\cdot \frac{p}{n_2}\right),\quad b= \al \frac{n_2}{n_1}\left( 1- \frac{p}{n_2}\right) + \left( 1- \frac{p}{n_1}\right). $$
Then we define the following parameters:
x= \frac{-b+ \sqrt{b^2 + 4a}}{2a},\quad y= \left[ x^{-2} - \frac{p}{n_1}\left( 1+\frac{p}{n_1}x\right)^{-2}\right]^{-1},\quad \omega= \al\frac{n_2}{n_1} \left( 1 + \al \frac{p}{n_1}x\right)^{-1}.
We have that
$$f(\al,n_1,n_2)= \frac{1 - 2\omega x + \omega^2 y}{ 1 - \frac{p}{n_2} \cdot \omega^2 y } +\oo(1)\quad \text{w.h.p.} $$
In the setting $\al=1$, both $\omega x$ and $\omega^2 y$ can be written in terms of only one parameter 
%$$f(\al,n_1,n_2)= \frac{\left(u^2 -\frac{p}{n_1}\right) \left(1- 2\frac{n_2}{n_1}  u^{-1}\right) + \frac{n_2^2}{n_1^2} }{ u^2 -\frac{p}{n_1}\left(1 + \frac{n_2}{n_1}\right) } +\oo(1)\quad \text{w.h.p.} $$
$$u: = x^{-1}\left(1+\frac{p}{n_1}x\right)=   \frac{ b+\sqrt{b^2+4a}}{2} +\frac{p}{n_1}   .$$
\fi
The proof of Theorem \ref{cor_MTL_loss} will be given in Appendix \ref{app_iso_cov} in the supplement \cite{MTL_suppl}. For the variance estimate in \eqref{Lvar_samplesize}, it is not necessary to assume the Gaussian distributions of the entries of $Z^{(1)}$ and $Z^{(2)}$. In fact, \eqref{Lvar_samplesize} is a special case of Theorem \ref{thm_main_RMT} below for the more general case with possibly different $\Sigma^{(1)}$ and $\Sigma^{(2)}$. On the other hand, the Gaussian assumption is needed in our current proof of the bias limit \eqref{Lbias_samplesize}. In the setting of Theorem \ref{cor_MTL_loss}, we can write  
L_{\bias} (a)&=\bv_a^\top (Z^{(1)})^\top Z^{(1)} \frac1{ \left[a^2(Z^{(1)})^\top Z^{(1)}+ (Z^{(2)})^\top Z^{(2)}  \right]^{2}}(Z^{(1)})^\top Z^{(1)} \bv_a,
where we denote $\bv(a):=(\Sigma^{(1)})^{1/2}\left(a\beta^{(1)}- a^2\beta^{(2)}\right)$. Using the rotational invariance of $(Z^{(1)})^\top Z^{(1)}$ and $(Z^{(2)})^\top Z^{(2)}$, we have that  
\begin{align}\label{Lbias_idea}
L_{\bias} (a)&\approx \|\bv_a\|^2 \frac1p\bigtr{ [(Z^{(1)})^\top Z^{(1)}]^2 \frac1{ \left[a^2(Z^{(1)})^\top Z^{(1)}+ (Z^{(2)})^\top Z^{(2)}\right]^{2}}  },
up to a small error. Notice that we can write \eqref{Lbias_idea} into a simpler form
\begin{align}\nonumber %\label{Lbias_idea2}
L_{\bias} (a)&\approx \|\bv_a\|^2 \left. \frac{\dd }{\dd x}\right|_{x=0}\frac1p\bigtr{  \frac1{ a^2(Z^{(1)})^\top Z^{(1)}+ x[(Z^{(1)})^\top Z^{(1)}]^2+(Z^{(2)})^\top Z^{(2)} }  }.
It is well-known that the empirical eigenvalue distributions (ESD) of $(Z^{(1)})^\top Z^{(1)}$ and $(Z^{(2)})^\top Z^{(2)}$ satisfy the famous Marchenko-Pastur (MP) law asymptotically \cite{MP}. From the MP law of $(Z^{(1)})^\top Z^{(1)}$, we can also derive the asymptotic ESD of $a^2(Z^{(1)})^\top Z^{(1)}+ x[(Z^{(1)})^\top Z^{(1)}]^2$ for any fixed $a\in \R$ and $x>0$. Due to the Gaussian assumption, $a^2(Z^{(1)})^\top Z^{(1)}+ x[(Z^{(1)})^\top Z^{(1)}]^2$ and $(Z^{(2)})^\top Z^{(2)}$ are asymptotically freely independent from each other. Hence the asymptotic ESD of $a^2(Z^{(1)})^\top Z^{(1)}+ x[(Z^{(1)})^\top Z^{(1)}]^2+(Z^{(2)})^\top Z^{(2)}$ is given by the free additive convolution (or free addition) of the asymptotic ESD of $a^2(Z^{(1)})^\top Z^{(1)}+ x[(Z^{(1)})^\top Z^{(1)}]^2$ and the MP law of $(Z^{(2)})^\top Z^{(2)}$. In particular, a sharp convergence estimate has been proved in \cite{BES_free1,BES_free2} for the free addition of two probability measure. We use that result to obtain a convergence estimate on 
$$\frac1p\bigtr{  \frac1{ a^2(Z^{(1)})^\top Z^{(1)}+ x[(Z^{(1)})^\top Z^{(1)}]^2+(Z^{(2)})^\top Z^{(2)} }  }.$$
Taking derivative with respect to $x$ at $x=0$ gives the exact asymptotic limit of $L_{\bias}(a)$.

We believe that the above argument can be extended to the case without Gaussian assumption. For example, instead of using the results in \cite{BES_free1,BES_free2} on free addition, we can use the sharp local laws on polynomials of random matrices in \cite{EKN_poly}. However, to apply the result in \cite{EKN_poly}, we need to check some hard technical conditions for our setting. Hence we do not pursue this direction in this paper.
%{\cob The idea of proof: proved using free probability; can also use polynomials of random matrices.}

Now we use the following example to illustrate the effect of varying sample sizes on the predication loss of the HPS estimator.

>>>>

\begin{abstract}
	\FY{rewrite the abstract} Hard parameter sharing for multi-task learning is widely used in empirical research despite the fact that its generalization properties have not been well established in many cases. This paper studies its generalization properties in a fundamental setting: How does hard parameter sharing work given multiple linear regression tasks? We develop new techniques and establish a number of new results in the high-dimensional setting, where the sample size and feature dimension increase at a fixed ratio. First, we show a sharp bias-variance decomposition of hard parameter sharing, given multiple tasks with the same features. Second, we characterize the asymptotic bias-variance limit for two tasks, even when they have arbitrarily different sample size ratios and covariate shifts. We also demonstrate that these limiting estimates for the empirical loss are incredibly accurate in moderate dimensions. Finally, we explain an intriguing phenomenon where increasing one task's sample size helps another task initially by reducing variance but hurts eventually due to increasing bias. This suggests progressively adding data for optimizing hard parameter sharing, and we validate its efficiency in text classification tasks.
\end{abstract}

@@

\FY{rewrite the introduction}  Hard parameter sharing (HPS) for multi-task learning is widely used in empirical research and goes back to the seminal work of \cite{C97}.
Recent work has revived interests in this approach because it improves performance and reduces the cost of collecting labeled data \cite{R17}.
It is generally applied by sharing the feature layers between all tasks while keeping an output layer for every task.
Often, hard parameter sharing offers two critical advantages if successfully applied.
First, it reduces model parameters since all tasks use the same feature space.
Second, it reduces the amount of labeled data needed from each task by augmenting the entire training dataset.

@@

One estimator that is sometimes employed in the literature is simply an average of the OLS estimators for two tasks \cite{bastani2020predicting}
\be\label{eq_est_ave}\hat \beta^{\AV}(\lambda)=\lambda \cdot \hat \beta_1^{\STL}+ (1-\lambda)\cdot \hat \beta_2^{\STL},\ee
where $\lambda\in [0,1]$ is an averaging parameter subject to choice in practice, and we recall that the OLS estimators are given by  
$$\hat \beta_i^{\STL}= [(X^{(i)})^\top X^{(i)}]^{-1}(X^{(i)})^\top Y^{(i)},\quad i=1,2.$$
The estimator in \eqref{eq_est_ave} is conventionally called an \emph{averaging estimator}. 
In particular, when $\lambda=0$ and $1$, we get the OLS estimators for tasks 2 and 1, respectively.

Another commonly used transfer learning estimator is defined as the minimizer of the total loss function \cite{bastani2020predicting, li2020transfer}:
$$\hat \beta^{\rm{TL}} := \argmin\left[ \norm{X^{(1)} \beta - Y^{(1)}}^2 + \norm{X^{(2)} \beta - Y^{(2)}}^2\right].$$
%\FY{Introduce some popular transfer learning estimators: total loss estimator, and model averaging estimator.}
In this paper, we consider a natural extension of the estimator $\hat \beta^{\rm{TL}} $, that is, the \emph{hard parameter sharing} (HPS) estimator, which has been a standard type of estimator in multi-task learning \FY{citations}. More precisely, we study the following HPS architecture: a shared feature representation layer $B\in\real^{p}$ for all datasets and a separate output layer $A_i \in \real$ for every dataset $i$. Then we study the following minimization problem:
\begin{align}\label{eq_tsl}
			f(A, B) = \norm{X^{(1)} B A_1 - Y^{(1)}}^2 + \norm{X^{(2)} B A_2 - Y^{(2)}}^2,
\end{align}
where we abbreviate $A = [A_1, A_2]$. Let $(\hat{A}, \hat{B})$ be the minimizer of $f(A, B)$. We define the hard parameter sharing (HPS) estimator for task $i$ as 
\be\label{def_HPS}\hat{\beta}_i^{\MTL} = \hat{B} \hat{A}_i,\quad i=1,2.\ee
Note that $\hat \beta^{\rm{TL}}$ is a special case of $\hat{\beta}_i^{\MTL}$ by setting $A_1=A_2=1$.

@@

In this paper, we measure the performance of $\hat{\beta}_2^{\MTL} (\hat a)$ through the out-of-sample predication loss (test error). Consider a test data point $(x,y)$ generated from the same model as task 2: $y= x^\top \beta^{(2)} + \ve$, where $x \in \R^p$ and $\ve\in \R$ are independent of $X^{(1)}$, $X^{(2)}$, $\varepsilon^{(1)}$ and $\varepsilon^{(2)}$, and only $x$ is observable. We want to use $x^\top \hat{\beta}_2^{\MTL} (\hat a) $ to predict $y$, and we measure the predication loss using the mean squared error
$$\exarg{x}{\left\|y-x^\top \hat{\beta}_2^{\MTL} (\hat a) \right\|^2}= \left\|(\Sigma^{(2)})^{1/2} \left(\hat{\beta}_2^{\MTL} (\hat a)  - \beta^{(2)}\right)\right\|^2 + \sigma^2.  $$ 
Since $\sigma^2$ is a constant that does not depend on the model, we ignore it and define the predication loss as
\be\label{HPS_loss}
L(\hat{\beta}_2^{\MTL}(\hat a)):= \left\|(\Sigma^{(2)})^{1/2} \left(\hat{\beta}_2^{\MTL} (\hat a)  - \beta^{(2)}\right)\right\|^2 .
 \ee
We will compare it to the out-of-sample predication losses of the baseline estimators, $L(\hat{\beta}_2^{\STL})$ and $L(\hat{\beta}^{\AV}(\lambda))$ for some properly chosen $\lambda\in [0,1]$.

%$L(\hat{\beta}_2^{\STL})$ of the single-task OLS estimator  as a baseline.

@@

 \begin{lemma}\label{lem_HPS_loss}
 Under Assumption \ref{assm_big1}, for any small constant $\e>0$ and large constant $C>0$, there exists a high probability event $\Xi$, on which the following estimates hold uniformly in $a\in \R$ and $\lambda\in [0,1]$: 
\begin{align}
L(\hat{\beta}_2^{\MTL}(a)) &=\left[ 1+ \OO(p^{-1/2+c})\right]\cdot \left[L_{\bias} (a) + L_{\var}(a)\right] \nonumber\\
&+ \OO \left[ p^{-C}\left( \|\beta^{(1)} \|^2  +  \|\beta^{(2)} \|^2\right)   \right], \label{L_HPS_simple}
\end{align}
\begin{align}
L(\hat{\beta}^{\AV}(\lambda) ) &= \left[ 1+ \OO(p^{-1/2+c})\right] \cdot \lambda^2 \left\| (\Sigma^{(2)})^{1/2}(\beta^{(1)}-\beta^{(2)})\right\|^2 \nonumber \\
&+  \left[ 1+ \OO(p^{-1/2+c})\right] \cdot \sigma^2   \lambda^2 \bigtr{\Sigma^{(2)} [(X^{(1)})^\top X^{(1)}]^{-1}  }\nonumber\\
& + \left[ 1+ \OO(p^{-1/2+c})\right] \cdot \sigma^2  (1-\lambda)^2 \bigtr{\Sigma^{(2)} [(X^{(2)})^\top X^{(2)}]^{-1}  }  . \label{L_AVE_simple}
\end{align}
and 
\begin{align}
L(\hat{\beta}_2^{\STL} ) &= \left[ 1+ \OO(p^{-1/2+c})\right] \cdot \sigma^2  \bigtr{\Sigma^{(2)} [(X^{(2)})^\top X^{(2)}]^{-1}  }  . \label{L_STL_simple}
\end{align}
%Moreover, there exists a constant $C_0>0$ such that with high probability,
%\be\label{loss_large}
%g(a) > g(0) \quad \text{for all $a$ such that }|a|\ge C_0.
%\ee
%with high probability for any small constant $\e>0$. 
\end{lemma}

@@

\begin{lemma}[Theorem 2.4 of \cite{isotropic} and Theorem 3.14 of \cite{DY}]\label{fact_tr}
	%Let $X  \in \real^{n\times p}$ be a random matrix that satisfies Assumption \ref{assume_rm}.
	%Let $\Sigma\in\real^{p\times p}$ denote the population covariance matrix of $X$.
	Under Assumption \ref{assm_big1}, we have that for any deterministic $p\times p$ matrix $\Sigma$,
		\[ \bigtr{\Sigma \frac1{(X^{(2)})^\top X^{(2)} }  } %= \bigtr{ \frac1{(Z^{(2)})^\top Z^{(2)} }  } 
		= \frac{1}{n_2 - p}\bigtr{\Sigma (\Sigma^{(2)})^{-1}} +  \OO\left( \frac{p^{\e}}{\sqrt{n_2 p}}\cdot \frac{p}{n_2}\|A\| \right) \]
		with high probability for any small constant $\e>0$.
\end{lemma}
 %The above result has a long history in random matrix theory.
If the entries of $Z^{(2)}$ are i.i.d. Gaussian, then this result follows from the classical result for %the mean of 
the inverse Wishart distribution \cite{anderson1958introduction}. For a general non-Gaussian $Z^{(2)}$, this result can be obtained using the well-known Stieltjes transform method (cf. Lemma 3.11 of \cite{bai2009spectral}). Here we have presented the results from \cite{isotropic,DY}, which give an almost sharp convergence rate. With Lemma \ref{fact_tr}, we obtain that w.h.p.,
\begin{align}
L(\hat{\beta}_2^{\STL} ) &= \left[ 1+ \OO(p^{-1/2+c})\right] \cdot \frac{p \sigma^2 }{n_2-p}   , \label{L_STL_simple01}
\end{align}
and
\begin{align}
L(\hat{\beta}^{\AV}(\lambda) ) &= \left[ 1+ \OO(p^{-1/2+c})\right] \cdot \lambda^2 \left\| (\Sigma^{(2)})^{1/2}(\beta^{(1)}-\beta^{(2)})\right\|^2 \nonumber \\
&+  \left[ 1+ \OO(p^{-1/2+c})\right] \cdot \sigma^2 \left[\frac{ \lambda^2}{n_1-p}\bigtr{\Sigma^{(2)}(\Sigma^{(1)})^{-1}} + \frac{(1-\lambda)^2p  }{n_2-p}\right] . \label{L_AVE_simple01}
 \end{align}

On the other hand, much less is known about the predication loss of the HPS estimator. In particular, its exact dependence on the model parameters, including the model bias, noise variance, sample sizes, and population covariance matrices, is not well-understood so far. %\FY{What is known so far about this topic?} 
In Section \ref{sec_HPS_loss}, we give a rigorous analysis of the bias term \eqref{Lbias} and variance term \eqref{Lvar} in two different settings: (i) independent $X^{(1)}$ and $X^{(2)}$ with different sample sizes and same population covariance matrices $\Sigma^{(1)}=\Sigma^{(2)}$; (ii) independent $X^{(1)}$ and $X^{(2)}$ with different sample sizes and different population covariance matrices. In case (i), we provide the exact asymptotics for both the bias and variance terms, together with almost sharp convergence rates. We will use these results to illustrate the effects of the \emph{bias-variance tradeoff} and \emph{varying sample sizes}. In case (ii), we will prove an exact asymptotic variance limit, and employ the result to show the intricate effect of the covariate shift. On the other hand, in case (ii), we will provide an estimate on the bias term, which becomes exact only when $n_1\gg p$. %Here case (ii) is actually more general than (i) 
%In particular, we will use our results to analyze the effect of the following factors: \emph{bias-variance tradeoff}, \emph{sample sizes}, and \emph{covariate shift}.   
In Section \ref{sec_same}, we will consider a simple extension to a multi-task learning setting with more than two tasks and  the same covariates for all tasks. We will give the asymptotic limit of the predication loss for the HPS estimator, which provides useful insights into the more rich multi-task learning theory. Due to length constraint, we will explore the multi-task learning setting in greater details in other works.

%{\cor discuss about the bias-variance trade-off}
%However, to the best of our knowledge, the asymptotic limit of $L(\hat{\beta}_2^{\MTL}(a))$ has never been calculated exactly in any high-dimensional setting.

\section{Predication loss for HPS estimator}\label{sec_HPS_loss}

@@

 In this subsection, we consider the setting where $X^{(1)}$ and $X^{(2)}$ are independent and have the same population covariance matrices $\Sigma^{(1)}=\Sigma^{(2)}$. However, the two tasks can have different sample sizes $n_1\ne n_2$. In this case, we can obtain the exact asymptotic limits of the bias term \eqref{Lbias} and the variance term \eqref{Lvar}. We will use these results to illustrate the effects of the bias-variance tradeoff and varying sample sizes on the performance of the $\MTL$ estimator.

@@

Under Assumption \ref{assm_big1}, suppose that $\Sigma^{(1)}=\Sigma^{(2)}$ and the entries of $Z^{(1)}$ and $Z^{(2)}$ are i.i.d. Gaussian random variables. Then for any small constant $\e>0$ and large constant $C>0$, there exists a high probability event $\Xi$, on which the following estimates hold uniformly in all $a\in \R$: 

@@

L_{\var}(a)= \sigma^2  \left[ \cal L_1(a)+ \OO\left({p^{c}}/{n_1}\right)\right]  ,

@@

L_{\bias}(a)&= \left[ \cal L_2(a)+  \OO\left(p^{-1/2+c}\right)\right] \cdot  \|\beta^{(1)}-a\beta^{(2)}\|^2 \nonumber\\

@@

\iffalse
 We need to state a result for Gaussian matrix ...... Consider 
$$f (\al,n_1,n_2)= \frac1p\tr\left[\frac{1}{ (X_1^\top X_1 + \al \cdot X_2^\top X_2)^2} (X_1^\top X_1)^2\right].$$
In our case, we have $\al=1$, but we can handle more general $\al$. We introduce two parameters:
$$a= \al \frac{n_2}{n_1} \left( \frac{p}{n_1} + \frac{p}{n_2}- \frac{p}{n_1 }\cdot \frac{p}{n_2}\right),\quad b= \al \frac{n_2}{n_1}\left( 1- \frac{p}{n_2}\right) + \left( 1- \frac{p}{n_1}\right). $$
Then we define the following parameters:
\begin{align*}
x= \frac{-b+ \sqrt{b^2 + 4a}}{2a},\quad y= \left[ x^{-2} - \frac{p}{n_1}\left( 1+\frac{p}{n_1}x\right)^{-2}\right]^{-1},\quad \omega= \al\frac{n_2}{n_1} \left( 1 + \al \frac{p}{n_1}x\right)^{-1}.
\end{align*}
We have that
$$f(\al,n_1,n_2)= \frac{1 - 2\omega x + \omega^2 y}{ 1 - \frac{p}{n_2} \cdot \omega^2 y } +\oo(1)\quad \text{w.h.p.} $$
In the setting $\al=1$, both $\omega x$ and $\omega^2 y$ can be written in terms of only one parameter 
%$$f(\al,n_1,n_2)= \frac{\left(u^2 -\frac{p}{n_1}\right) \left(1- 2\frac{n_2}{n_1}  u^{-1}\right) + \frac{n_2^2}{n_1^2} }{ u^2 -\frac{p}{n_1}\left(1 + \frac{n_2}{n_1}\right) } +\oo(1)\quad \text{w.h.p.} $$
%where 
$$u: = x^{-1}\left(1+\frac{p}{n_1}x\right)=   \frac{ b+\sqrt{b^2+4a}}{2} +\frac{p}{n_1}   .$$
\fi

We believe the convergence rates $p^c/n_1$ and $p^{-1/2+c}$ in \eqref{Lvar_samplesize} and \eqref{Lbias_samplesize} are both sharp up to the $p^c$ factor. The proof of Theorem \ref{cor_MTL_loss} will be given in Appendix \ref{app_iso_cov} in the supplement \cite{MTL_suppl}. For the variance estimate in \eqref{Lvar_samplesize}, it is not necessary to assume the Gaussian distributions of the $Z^{(1)}$ and $Z^{(2)}$ entries. In fact, \eqref{Lvar_samplesize} is a special case of Theorem \ref{thm_main_RMT} below for the more general case with possibly different $\Sigma^{(1)}$ and $\Sigma^{(2)}$. On the other hand, the Gaussian assumption is needed in our current proof of the bias limit \eqref{Lbias_samplesize}. In the setting of Theorem \ref{cor_MTL_loss}, we can write  
\begin{align*}
L_{\bias} (a)&=\bv_a^\top (Z^{(1)})^\top Z^{(1)} \frac1{ \left[a^2(Z^{(1)})^\top Z^{(1)}+ (Z^{(2)})^\top Z^{(2)}  \right]^{2}}(Z^{(1)})^\top Z^{(1)} \bv_a,
\end{align*}
where we abbreviate $\bv(a):=(\Sigma^{(1)})^{1/2}\left(a\beta^{(1)}- a^2\beta^{(2)}\right)$. By the rotational invariance of $(Z^{(1)})^\top Z^{(1)}$ and $(Z^{(2)})^\top Z^{(2)}$, we have that  
\begin{align}\label{Lbias_idea}
L_{\bias} (a)&\approx \|\bv_a\|^2 \frac1p\bigtr{ [(Z^{(1)})^\top Z^{(1)}]^2 \frac1{ \left[a^2(Z^{(1)})^\top Z^{(1)}+ (Z^{(2)})^\top Z^{(2)}\right]^{2}}  },
\end{align}
up to a small error. Notice that we can write \eqref{Lbias_idea} into a simpler form
\begin{align}\nonumber %\label{Lbias_idea2}
L_{\bias} (a)&\approx \|\bv_a\|^2 \left. \frac{\dd }{\dd x}\right|_{x=0}\frac1p\bigtr{  \frac1{ a^2(Z^{(1)})^\top Z^{(1)}+ x[(Z^{(1)})^\top Z^{(1)}]^2+(Z^{(2)})^\top Z^{(2)} }  }.
\end{align}
It is well-known that the empirical spectral distributions (ESD) of $(Z^{(i)})^\top Z^{(i)}$, $i=1,2$, satisfy the famous Marchenko-Pastur (MP) law asymptotically \cite{MP}. From the MP law of $(Z^{(1)})^\top Z^{(1)}$, we can also derive the asymptotic ESD of $a^2(Z^{(1)})^\top Z^{(1)}+ x[(Z^{(1)})^\top Z^{(1)}]^2$ for any fixed $a\in \R$ and $x>0$. Due to the rotational invariance of multivariate Gaussian distributions, 
%$a^2(Z^{(1)})^\top Z^{(1)}+ x[(Z^{(1)})^\top Z^{(1)}]^2$ and $(Z^{(2)})^\top Z^{(2)}$ are asymptotically freely independent from each other \cite{nica2006lectures}. Hence 
the asymptotic ESD of $a^2(Z^{(1)})^\top Z^{(1)}+ x[(Z^{(1)})^\top Z^{(1)}]^2+(Z^{(2)})^\top Z^{(2)}$ is given by the free additive convolution (or free addition) of the asymptotic ESD of $a^2(Z^{(1)})^\top Z^{(1)}+ x[(Z^{(1)})^\top Z^{(1)}]^2$ and the MP law of $(Z^{(2)})^\top Z^{(2)}$\cite{nica2006lectures}. In particular, a sharp convergence estimate has been proved in \cite{BES_free1,BES_free2} for the free addition of two probability measures. We use that result to obtain a convergence estimate on 
$$\frac1p\bigtr{  \frac1{ a^2(Z^{(1)})^\top Z^{(1)}+ x[(Z^{(1)})^\top Z^{(1)}]^2+(Z^{(2)})^\top Z^{(2)} }  }.$$
Taking derivative with respect to $x$ at $x=0$ gives the exact asymptotic limit of $L_{\bias}(a)$. We refer the reader to Appendix \ref{app_iso_cov} of the supplement \cite{MTL_suppl} for more details.

We believe that the above argument can be extended to the case without the Gaussian assumption. For example, instead of using the results in \cite{BES_free1,BES_free2} on free addition, we can use the sharp local laws on polynomials of random matrices in \cite{EKN_poly}. However, to apply the result in \cite{EKN_poly}, we need to check some difficult technical regularity conditions for our setting, which is not the focus of this work. Hence we do not pursue this direction in this paper.
%{\cob The idea of proof: proved using free probability; can also use polynomials of random matrices.}

%Now we use the following example to illustrate the effect of varying sample sizes on the predication loss of the HPS estimator.

To describe quantitively the bias-variance tradeoff phenomenon and the effect of varying sample sizes on the predication loss, we consider a \emph{random-effect model}, which has been studied for  single-task linear regression and ridge regression (see e.g. \cite{dobriban2020wonder,dobriban2018high}).
